Automatic discovery of cell types and microcircuitry from neural connectomics

### 2014-12-10 docs/processor.tex
updates to experiment runners

%\documentclass{nature}
\documentclass{article}
\usepackage[american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage{todonotes}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
%\usepackage{algorithm}
%\usepackage{algorithmic}
\usepackage{amsmath}
%\usepackage{subcaption}
%\usepackage{multicol}
\usepackage{csquotes}
\usepackage{pdfpages}
\usepackage{placeins}

\renewcommand{\vec}[1]{\mathbf{#1}}

%\renewcommand{\includegraphics}[2][]{\fbox{}}

\addbibresource{connectomics motif.bib}
\addbibresource{connectomics motif-relational models.bib}

\DeclareGraphicsRule{.ai}{pdf}{.ai}{}

\title{Automatic discovery of cell types and microcircuitry from neural connectomics}
\author{Eric Jonas$^1$, Konrad Kording$^{2, 3,4}$}

\begin{document}
\maketitle

\begin{small}
\begin{enumerate}
  \item Electrical Engineering and Computer Science, University of California, Berkeley
  \item Department of Physical Medicine and Rehabilitation,
    Northwestern University and Rehabilitation Institute of Chicago,
    Chicago, Illinois
  \item Department of Physiology, Northwestern University, Chicago, Illinois
  \item Department of Applied Mathematics, Northwestern University, Chicago, Illinois
\end{enumerate}
\end{small}

\begin{abstract}
  Neural connectomics has begun producing massive amounts of data,
  necessitating new analysis methods to discover the biological and
  computational structure. It has long been assumed that discovering
  neuron types and their relation to microcircuitry is crucial to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. It combines the information
  traditionally used by biologists, including connectivity, cell body
  location and the spatial distribution of synapses.  We show that the
  approach recovers known neuron types in the retina and enables
  predictions of connectivity, better than simpler algorithms. It also
  can reveal interesting structure in the nervous system of
  c. elegans, and automatically discovers the structure of a
  microprocessor.  Our approach extracts structural meaning from
  connectomics, enabling new approaches of automatically deriving
  anatomical insights from these emerging datasets.
\end{abstract}

\section*{Introduction}
Emerging connectomics techniques \autocite{Morgan2013,Zador2012}
promise to quantify the location and connectivity of each neuron
within a tissue volume. These massive datasets will far exceed the
capacity of neuroanatomists to manually trace small circuits, thus
necessitating computational, quantitative, and automatic methods for
understanding neural circuit structure.  The impact of this kind of
high-throughput transition has been seen before -- rise of sequencing
techniques necessitated the development of novel computational methods
to understand genomic structure, ushering in  bioinformatics
as an independent discipline \autocite{Koboldt2013}.

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined, type-dependent local
connectivity patterns (microcircuits) has been prominent in many
areas, from sensory (e.g. retina, \autocite{Masland2001} to processing
(e.g. neocortex \autocite{Mountcastle1997}) to movement (e.g. spinal
cord) \autocite{Grillner2005}. These sorts of repeated computing
patterns are a common feature of computing systems, even arising in
human-made computing circuits. It remains an important challenge to
develop algorithms to use anatomical data, e.g. connectomics, to
automatically back out underlying microcircuitry.

The discovery of structure is a crucial aspect of network
science. Early approaches focused on global graph properties, such as
the types of scaling present in the network \autocite
{WattsStrogatz1998}.  While this approach leads to an understanding
of the global network, more recent work aims at identifying very small-scale
repeat patterns, or “motifs” in networks\autocite{Milo2002}. These motifs
are defined not between different node types, but rather represent repeated
patterns of topology.

The discovery of structure in probabilistic graphs is a crucial aspect
in machine learning. Commonly used algorithms include
community-based-detection methods \autocite{Girvan2002}, and
stochastic block models \autocite{Nowicki2001}.  While these
approaches can incorporate the probabilistic nature of neural
connections \autocite{Hill2012} they do not incorporate the additional
richer structure present in connectomics data -- the location of cell
bodies, the spatial distribution of synapses, and the distances
between neurons. Of particular importance is that the probability of
connections has a strong spatial component, a factor that is hard to
reconcile with many other methods. A model attempting to fully capture
the variation in the nervous system should take into account the broad
set of available features.

When it comes to neuroscience and other computing systems, we expect
patterns of connectivity much more complex than traditional motifs,
exhibiting a strong spatial dependence arising from the complex
genetic, chemical, and activity-based neural development processes.

To address these challenges, here we describe a Bayesian
nonparametric model that can discover circuit structure automatically
from connectomics data: the cell types, their spatial patterns of
interconnection, and the locations of somata and synapses. We show
that by incorporating this additional information, our model both
accurately predicts the connection as well as agrees
with human neuroanatomists as to the identification of cell types.

We primarily focus on the recently-released mouse retina connectome
\autocite{Helmstaedter2013}, but additionally examine the c. elegans
connectome \autocite{White1986}, and then ``connectome'' of a
classical microprocessor \autocite{James2010}. Comparing the cell
types discovered by the algorithms with those obtained manually by
human anatomists reveals a high degree of agreement. We present a
scalable approach to infer microcircuitry from connectomics data.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{fig1b.ai}}
  \caption{An overview of our method. a. Input data is a connectivity
    matrix indicating whether cell $i$ (row) synapses on cell $j$
    (column), a distance matrix between cell soma, b. the depth of the
    cell soma, and the c. depth profile of the cells synaptic
    connectivity. d. Our algorithm discovers hidden cell types in this
    connectivity data by assuming all cells of a type share a
    distance-dependent connectivity profile, similar depth, and a
    similar synpatic density profile, with cells of other types.  This
    results in a clustering of the cells by those hidden
    types. d. shows the cell connectivity matrix with cells of the
    same type grouped together. e.) shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. f. shows a connectivity profile with virtually no
    probability of connectivity. We can plot this these types g.) to
    show the relationship and spatial connections, and how probable
    various types are to connect to one another. h. The connectivity
    between our discovered types changes as a function of distance. }

\label{fig:overview}
\end{figure}

\section{Results}

We start with the generic assumption that cell types exist -- a cell
is a member of a single, unobserved type -- and these types affect
soma depth, distribution of synapses, as well as a cell type and
distance dependent connection probability. From this assumption we
develop a generative Bayesian model that estimates the underlying cell
types and how they connect.  We start with a model for connectivity,
the infinite stochastic block model (iSBM)\autocite{Kemp2006a,Xu2006},
which has been shown to meaningfully cluster connection graphs while
learning the number of hidden groups, or types. We extend this
approach by adding distance dependence to model salient aspects of
microcircuitry. We additionally model cell body depth as a mixture of
Gaussians and each cell’s synapse density profile as mixture of
mixtures of Gaussians.

For example, retinal ganglion cells may synapse on nearby, but not far
away, amacrine cells, with ganglion cells being superficial and having
a broad distribution of synapses.  This procedure automatically learns
the number of cell types, which cells belong to which types, along
with the distance and type dependent connectivity probabilities (see
fig~\ref{fig:overview}). We apply our algorithm to datasets from mouse
retina, c-elegans and a historical microprocessor. Anatomists classify
cells based on many features, giving us a meaningful baseline to
compare against.

Before we can use a probabilistic model to analyze real data we first
need to assess if the model works properly. We thus simulate data for
which we know the correct structure and comparing the estimated
structure based on the algorithm (see methods) with the one we used
for simulation. We find that the model does a good job of recovering
the correct number of cell types, (fig~\ref{fig:synthetic}a), the the
cell identities (fig~\ref{fig:synthetic}b), and the spatial extent of
each type (fig~\ref{fig:synthetic}c).  Indeed, it does far better than
the existing infinite stochastic block model that assumes that there
is no distance dependence to the connectivity. The model converges
relatively quickly to an estimate of the most probable values for the
cell types, which is enabled by using a combination of simulated
annealing and parallelized Markov chain Monte Carlo (see methods for
details). It appears that the model behaves the way it should and
allows application to relatively large datasets.

\begin{figure}
  \centering 
  \centerline{\includegraphics{synthetic.ai}}
  \caption{Correct recovery of true numbers of hidden types in
    synthetic data when incorporating spatial information. a.) The
    infinite stochastic block model (which only uses connectivity
    information) over-estimates the number of classes as it fails to
    take distance into account, whereas our modeling of the
    combination of distance and connectivity finds close to the true
    number of classes. b.) ARI, a measure of ``correct clustering'',
    for different true class counts, between our model and a
    traditional iSBM. c.) For synthetic data, type discovery based
    only on connectivity are tiny spatially-localized groups -- in
    contrast, our model recovers the true spatial extent of the
    underlying types.  }
\label{fig:synthetic}
\end{figure}

\subsubsection{Mouse Retina}

\begin{figure}
  \centering 
  \centerline{\includegraphics{mouseretina.ai}}
  \caption{Discovering cell classes in the mouse retina connectome.
    a.) Input connectivity data for 950 cells for which soma positions
    were known. b. clustered connectivity matrix. c.) connectivity
    diagram showing our clusters, as well as the cell depth and
    anatomist-labeled cell type of each cell. Our model shows moderate
    agreement with the cell types traditionally identified by
    anatomist laminar-specific connectivity patterns. 
    d.) Connectivity between our clusters as a function
    of distance -- the cluster consisting primarily of retinal
    ganglion cells (lower left-center) exhibits the expected near and
    far connectivity. e.)  The spatial distribution of our cell types
    -- each cell type tesselates space.}
\label{fig:mouseretina}
\end{figure}

The mouse retina \autocite{Masland2001} is a neural circuit which we
expect to have connectivity patterns that are well approximated by our
generative model. It is known that there are multiple classes of cells
that can be broadly grouped into: ganglion cells that transmit
information to the rest of the brain, bipolar cells that connect
between different cells, and amacrine cells that feed into the
ganglion cells. Recent research \autocite{Helmstaedter2013} has
produced a large dataset containing both the types of cells from
orthogonal approaches, and also the connectivity matrix between all
reconstructed cells.

The algorithm took less than 2 hours to perform inference, dividing
neurons into a set of cell types (fig~\ref{fig:mouseretina}c, each
wedge is a type). For each pair of neurons there is a specific
distance dependent connection probability
(fig~\ref{fig:mouseretina}b,c,d), which is well approximated by the
model fit. Moreover, each type of cell is rather isotropically
distributed across space (fig~\ref{fig:mouseretina}e) as should be
expected for true cell types.

Comparing the results of the algorithm to other information sources
allows evaluating the quality of the type determination. Our types
closely reflect the (anatomist-determined) segmentation of cells into
retinal ganglion, narrow amacrine, medium/wide amacrine, and bipolar
cells (fig~\ref{fig:mouseretina}c, outermost ring). We find that the
types we find tend to reflect the known laminar distribution in the
retina (fig~\ref{fig:mouseretina}c, middle ring).

The algorithm yields a separation of neurons into a smaller number of
types than the fully granular listing of 71 types found by the
original authors of the paper, although is still highly correlated
with those finer type distinctions (see
supp~\ref{supp:mouseretina}. It is our expectation that, with larger
datasets, even closer agreement would be found.

Our fully Bayesian model produces a distribution over probable
clusterings.  Figure~\ref{fig:mouseretina_compare} shows this
posterior distribution as a cell-cell coassignment matrix, sorted to
find maximum block structure. Each large, dark block represents a
collection of cells believed with strong probability to be of the same
type. When we plot (fig~\ref{fig:mouseretina_compare}b) the
anatomist-derived cell types along the left, we can see that each
block consists of a roughly-homogeneous collection of types.

We evaluate our model along three sets of parameters
(Fig.\ref{fig:mouseretina_compare}): how closely does our clustering agree
with neuroanatomists' knowledge?  Given two cells, how accurately can
our model predict the link between them? And how closely does the
spatial extent (within a layer) of our identified types agree with the
known neuroanatomists.

As a first measure we compare link prediction accuracy across the
methods (Fig.\ref{fig:mouseretina_compare} B AUC, red).  We find that
given the dataset many techniques allow for good link-predictive
accuracy. All the methods allow decent link prediction with an AUC in
the .9 range. However, our algorithm clearly outperforms the simple
statistical models that only use connectivity.

As a second measure we compare link prediction accuracy across the
methods (Fig.\ref{fig:mouseretina_compare} B ARI, blue). We find that our
algorithm far outperforms the controls. We also find that when it is
based on more of the same information used by anatomists use then it
gets better at agreeing with these anatomists. In particular, using
connectivity, distance, synapse distribution and soma depth leads to
the highest ARI. When using the available information the algorithm
produces a good fit to human anatomist judgments.

Finally we look at the spatial extent of the discovered types both
within a layer and between layers (Fig.\ref{fig:mouseretina_compare}
C). We see that, in the absence of distance information, mere
connectivity information results in types which only span a small
region of space -- essentially “local cliques”. Incorporation of
distance information results in types which span the entire extent of
the layer. The depth variance of all models continues to be
substantially larger than that predicted by human anatomists -- future
directions of work include attempting to more strongly encode this
prior belief of laminarity.

\begin{figure}
  \centering 
  \centerline{\includegraphics{comparison.ai}}
  \caption{a. Cell-cell assignment matrix reflecting the posterior
    distribution of clusterings found by our model. Large contiguous
    black areas are cells that were considered the same type in the
    majority of samples. b. The cells and their true types sorted to
    agree with the coassignment matrix. c. A comparison of the
    predictive accuracy (Area under the curve) for hand-labeled
    anatomical data, versus inclusion of additional sources of
    information. d. ) The spatial extent (in depth and area) of the
    types identified by humans and our various algorithmic
    approaches.}

\label{fig:mouseretina_compare}
\end{figure}

% from experiments/mouseretina/spatial_var.txt

% Relational Model: mean = 14.586508 std=8.136334
% Spatial-Relational Model: mean = 34.540355 std=1.069483
% Truth (fine): mean = 34.348443 std=7.567564
% Truth (coarse): mean = 35.276636 std=2.372951

\subsection{C. elegans}

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons developmentally determined,
leading to early measurement of the connectome. Unlike the retina,
only the motor neurons in c. elegans exhibit regular distribution in
space -- along the body axis. Most interneurons are
concentrated in various ganglia that project throughout the entire
animal, and the sensory neurons are primarily located in a small
number of anterior ganglia.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=7in]{otherdata.ai}}
  \caption{Discovering connectivity and type in both
    \textit{c. elegans} and the MOS 6502 microprocessor.  a.) Initial
    \textit{c. elegans} cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). d.) The clusters and connectivity versus
    distance for connections between Gate and C1, Gate and C2, and C1
    and C2 terminals on a transistor. Red and teal types have a
    terminal pulled down to ground and mostly function as
    inverters. Purple class are clocked, stateful transistors, orange
    control the ALU and yellow control the special data bus. e.) The
    spatial distribution of the transistors in each cluster.}
  \label{fig:otherdata}
\end{figure}

Using both the chemical and electrical connectivity (see methods), we
determined the underlying clusters explained by connectivity and
distance (fig~\ref{fig:otherdata}a). A superficial inspection of the results shows
clustering into groups consisting roughly homogeneously of motor
neurons, sensory neurons, and interneurons. Closer examination reveals
agreement with the classifications originally outlined by Brenner in
1986.

We identify cell types that reflect the known motor/non-motor neuron
classes, even though this system lacks the strong repeat
microcircuitry our model was designed for.  Motor-neuron types AS, DA,
and VA, all exclusively postsynaptic, are identified as a common type,
as are motor-neuron types VD and DD. Traditional types VC, DB, and VB
also mostly share a cluster. Various head motor neurons, including
types SMD and RMD, are clustered together. Interneurons with known
anatomically-distinct connectivity patterns, such as AVA (2 cells),
are clustered into pure types.The algorithm even correctly places the
single-cell types DVB and DVC by themselves.

Note our clustering does not perfectly reflect known divisions --
several combinations of head and sensory neurons are combined, and a
difficult-to-explain group of mostly VB and DB motor neuron types,
with VC split between various groups. Our identified cell types thus
reflect a ``coarsening'' of known types, based entirely on
connectivity and distance information, even when the organism exhibits
substantially less spatial regularity than the retina.

\subsection{Microprocessor}

To show the applicability of our method to other connectome-style
datasets, we obtained the spatial location and interconnectivity of
the transistors in a classic microprocessor, the MOS Technology 6502
(used in the Apple II) \autocite{James2010}. Computer architects use
common patterns of transistors when designing circuits, with each
transistor having a ``type'' in the circuit. We identified a region of
the processor with complex but known structure containing the primary
8-bit registers X, Y, and S (fig~\ref{fig:otherdata}).

Our algorithm identifies areas of spatial homogeneity that mirror the
known structure in the underlying architectural circuit, segmenting
transistor types recognizable to computer architects. Using the
original schematics, we see that one identified type contains the
``clocked'' transistors, which retain digital state. Two other types
contain transistors with pins C1 or C2 connected to ground, mostly
serving as inverters.  An additional identified type controls the
behavior of the three registers of interest (X, Y, and S) with respect
to the SB data bus, either allowing them to latch or drive data from
the bus. The repeat patterns of spatial connectivity are visible in
figure~\ref{fig:otherdata}a, showing the man-made horizontal and
vertical layout of the same types of transistors.

\section{Discussion}
We have presented a machine learning technique that allows cell types
and microcircuitry to be discovered from connectomics data.  We have
shown its applicability to a newly published connectomics dataset from
the mouse retina. When compared to existing methods, we show how the
incorporation of all of this data yields results that combine both
high link-prediction accuracy and high agreement with human
anatomists.  I also reveals interesting structure when applied to the
classical connectivity of the worm c. elegans, and the connectivity of
a historical man-made microprocessor. We have found that combining the
available data types allows us to discover cell types and
microcircuitry that were known to exist in the systems based on
decades of previous research and allows good prediction of
connectivity.

For our probabilistic models, no known solution exists to
exactly find the most probable parsing of the neurons into cell-types
and connectivity patterns. We employ a collection of Markov-chain
Monte carlo techniques (see Methods) but while different
initializations converge to similar ultimate values, we can never
realistically obtain the global optimum. There are a broad range of
techniques that may offer good approximations to the global optimum
and future work could adapt them to
find more precise solutions to our problem.

For our probabilistic model, inference becomes slower as the amount of
data increases. Our algorithm required several hours for 1000
neurons. Scaling this class of probabilistic model is an active area
of research, and recent results in both variational methods
\autocite{Hoffman2013} and spectral learning \autocite{Anandkumar2012}
and future work could adapte them to find faster approximate solutions
to our problem.

Larger datasets will allow algorithms to distinguish more distinct
types and we expect closer agreement with existing anatomical
knowledge as more data become available.  Moreover, in general, for
such problems precision increases with the size of the dataset and the
cells that we have are not sufficient to statistically distinguish all
the cell types known in anatomy (such as the $\sim 70$ in the
retina). Still, using only connectivity and distance it is possible to
meaningfully divide neurons into types.

Our small collection of hand-selected distance-dependent likelihood
functions are clearly non-exhaustive, and assume monotonicity
of connectivity probability -- for a given class, closer cells
are never less-likely to connect. This is known to be insufficient
for various neural systems. Future models could incorporate
a wider variety of likelihood functions, or even learn the global
functional form from the data.

There exist a range of previous approaches to the discovery of neural
microcircuitry\autocite{Mountcastle1957, Douglas1991, Bartho2004,
  Freund1998}.  These generally involve a great deal of manual labor
and ad-hoc determination of what constitutes a “type” of cell -- to
this day there are disagreements in the literature as to the “true”
types in the mammalian retina. Much as phylogenomics has changed our
understanding of animal ontologies, modern large scale data will allow
the efficient unbiased discovery of cell types and circuits. The sheer
amount of available data demands the introduction of algorithmic
approaches.

The development of automatic identification and quantification of cell
type may also provide a new computational phenotype for quantifying
the effect of disease, genetic interventions, and
developmentally-experienced neural activity. Our method can in
principle identify neuron-types across non-connected graphs,
e.g. across animals. For example, the types of neurons in one animal
can be associated with the types of neurons in another animal, in the
same way as this is already possible through molecular marker
\autocite{Brown2009}. This could be particularly important if cell
types appear that are due to properties of the stimuli and experience
as opposed to just the molecular properties of cells, such as color
and orientation selective types in primary visual cortex
\autocite{Sincich2005,Lennie2005}. This would allow comparative
quantitative anatomy across animals, and aid the search for the
ultimate causes of connectivity.

Our model combines connectivity, cellular, and synaptic properties,
and suggests the way towards combining even richer data. Distinct cell
types differ in morphology, connectivity, transcriptomics, relation to
behavior or stimuli and many other ways. Algorithms combining this
data and type type information may allow us to synthesize all the
available information from one experiment or even across experiments
into a joint model of brain structure and function.

Eventually, algorithms will have to become a central tool for
anatomists, as it will progressively become impossible for humans to
parse the huge datasets. This transition may follow a similar
transition to the one underway in molecular biology (with gene-finding
algorithms) and evolutionary biology with (computational
phylogenetics).  In fact, it is known that there are significant
disagreements across human anatomists and computational approaches
promise to make the debates more data driven.

\subsection{Methods Summary}

For the basic link-distance model, we take as input a connectivity
matrix $R$ defining the connections between cell $e_i$ and $e_j$, as
well as a distance function $d(e_i, e_j)$ representing a (physical)
distance between adjacent cells.  See
the supplemental material for extension to multiple connectivity
matrices. We assume there exist an unknown number $K$ of latent
(unobserved) cell types, $k \in \{1, 2, 3, \dots, K\}$, and that each
cell $e_i$ belongs to a single cell type. We indicate a cell $e_i$ is
of type $k$ using the assignment vector $\vec(c)$, so $c_i = k$. The
observed connectivity between two cells $R(e_i, e_j)$ then depends
only on their latent type and their distance through a link function
$f(\cdot, d(e_i, e_j))$. We assume $f$ is parameterized based on the
latent type, $c_i=m$ and $c_j=n$, via a parameter$\eta_{mn}$, as well
as a set of global hyper parameters $\theta$, such that the link
function is $f(d(e_i, e_j) | \eta_{mn}, \theta)$.

We then jointly infer the maximum a posteriori (MAP) estimate of the
class assignment vector $\vec(c) = \{c_i\}$, the parameter matrix
$\eta_{mn}$, and the global model hyperparameters $\theta$ :

\begin{equation}
  p(\vec{c}, \eta, \theta | R ) \propto \prod_{i, j} p(R(e_i, e_j) | f(d(e_i, e_j) | \eta_{c_ic_j}), \theta) \prod_{m, n} p(\eta_{mn} | \theta)  p(\theta) p(\vec{c} | \alpha) p(\alpha) p(\theta)
\end{equation}

For the retina data, we then extend the model with the additional
features indicated. Cell soma depth is modeled as a
cell-type-dependent Gaussian with latent (unknown) per-type mean and
variance. Similarly, each cell has some number $N_i$ of synapses, 
each of which is drawn from a cell-type-specific density profile
with up to three modes.

Inference is performed in three steps via composible transition 
kernels -- one for structural, one for per-type parameters, and
one kernel for global parameters and hyperparameters. Details
of data preprocessing, inference parameters, and runtime can
be found in the Methods section.

\printbibliography

\begin{itemize}

 \item \textbf{Acknowledgments} We thank Josh Vogelstein for discussions and reading of the manuscript, Finale Doshi-Velez for early discussions on the model, and Erica Peterson, and Jonathan Glidden for extensive manuscript review. Funding for compute time was provided by Amazon Web Services ``AWS in Education'' grants. 
\item \textbf{Author Contributions} KK and EJ developed model. EJ derived inference, implemented code, tested, and ran experiments. KK and EJ wrote manuscript text and solicited feedback. 
 \item \textbf{Competing Interests} The authors declare that they have no
competing financial interests.
 \item \textbf{Correspondence} Correspondence and requests for materials
should be addressed to E.J.~(email: jonas@eecs.berkeley.edu).
\end{itemize}

\newpage
\section{Methods }

\subsection{Probabilistic Model}

Our model is a extension of the iSBM
\autocite{Kemp2006a,Xu2006} to incorporate spatial relations between entities,
inspired by attempts to extend these models with arbitrary
discriminative functions\autocite{Murphy2012}.

We take as input a connectivity matrix $R$ defining the connections
between cell $e_i$ and $e_j$, as well as a distance function $d(e_i,
e_j)$ representing a (physical) distance between adjacent cells. See
the supplemental material for extension to multiple connectivity
matrices. We assume there exist an unknown number $K$ of latent
(unobserved) cell types, $k \in \{1, 2, 3, \dots, K\}$, and that each
cell $e_i$ belongs to a single cell type. We indicate a cell $e_i$ is
of type $k$ using the assignment vector $\vec(c)$, so $c_i = k$. The
observed connectivity between two cells $R(e_i, e_j)$ then depends
only on their latent type and their distance through a link function
$f(\cdot, d(e_i, e_j))$. We assume $f$ is parameterized based on the
latent type, $c_i=m$ and $c_j=n$, via a parameter$\eta_{mn}$, as well
as a set of global hyper parameters $\theta$, such that the link
function is $f(d(e_i, e_j) | \eta_{mn}, \theta)$.

We then jointly infer the maximum a posteriori (MAP) estimate of the
class assignment vector $\vec(c) = \{c_i\}$, the parameter matrix $\eta_{mn}$, and
the global model hyperparameters $\theta$ :

\begin{equation}
  p(\vec{c}, \eta, \theta | R ) \propto \prod_{i, j} p(R(e_i, e_j) | f(d(e_i, e_j) | \eta_{c_ic_j}), \theta) \prod_{m, n} p(\eta_{mn} | \theta)  p(\theta) p(\vec{c} | \alpha) p(\alpha) p(\theta)
\end{equation}

% Methods section 3000 words
% methods summary does not appear in the online text
% methods section can have no figures or tables [grrr]

We describe the spatial ``Logistic-distance Bernoulli''  function here,
and others in the supplemental material.

% fixme class component, NOT class
The ``logistic-distance Bernoulli'' spatial model assumes that, if cell
$e_i$ is of type $m$ and cell $e_j$ is of type $n$, then $\eta_{mn}
= (\mu_{mn}, \lambda_{mn})$, and the probability that two cells $e_i$
and $e_j$ are connected is given by
\begin{eqnarray}
p^* &=& \frac{1.0}{1 + \exp \frac{d(e_i, e_j) - \mu_{mn}}{\lambda_{mn}}}\\
p &= & p^* \cdot (p_{max} - p_{min}) + p_{min}
\end{eqnarray}
where $p_{max}$ and $p_{min}$ are global per-graph parameters.

We place an exponential priors on the latent parameters:
\begin{eqnarray}
 \mu_{mn} \sim \exp(\mu | \mu^{hp}) \\
\lambda_{mn} \sim \exp(\lambda | \lambda^{hp})
\end{eqnarray}

using  $\lambda^{hp}$ and $\mu^{hp}$ as global per-graph hyperparameters.

We use a Dirichlet-process prior on class assignments, which allows
the number of classs to be determined automatically. In brief, for $N$
total cells, the probability of a cell belonging to a class is
proportional to the number of datapoints already in that class, $N_k$,
such that $p(c_i = k) \propto \frac{m_k}{N + \alpha}$ and the
probability of the cell belonging to a new class $k'$ is $p(c_i = k')
\propto \frac{\alpha}{N + \alpha}$. $\alpha$ is the global
concentration parameter -- larger values of $\alpha$ make the model
more likely to propose new classes. We grid the parameter $\alpha$ and
allow the best value to be learned from the data.

Where we model cell depth, we assume that 
each cell type has a typical depth, and thus a Gaussian distribution
of $s_i$. We assume $s_i \sim N(\mu^{(s)}_k,
\sigma^{2(s)}_k)$, where the $(s)$ superscript indicates
these model parameters are associated with the soma-depth 
portion of our model.  We use a conjugate prior for $(\mu^{(s)}_k, \sigma^{2(s)}_k)$ with
$\mu^{(s)}_k \sim N(\mu^{(s)}_{hp}, \sigma^{2(s)}_k/ \kappa^{(s)}_{hp})$ and $\sigma^{2(s)}_k \sim
\chi^{-1}(\sigma^{2(s)}_{hp}, \nu^{(s)}_{hp}$. The use of conjugacy simplifies inference while allowing for each cell-type to have its own depth mean and distribution.

Where we model synapse depth profile, we assume that each cell
type has a characteristic depth distribution of synaptic contact points, and thus
a mixture of Gaussians distribution over cell $i$’s $N_i$ contact points, $\vec{g^i}$.
We do this by assuming the $g^i_j$
are drawn from an $M=3$-component mixture of Gaussians. Thus associated with each cell type $k$ is a vector
of $M$ Gaussian means $(\mu^g_{k,1}, \cdots, \mu^g_{k, M})$, and a
mixture vector $\pi_k$.  This representation can thus model depth distributions of contact points that have up to three modes, an assumption that is well matched in the bulk of anatomical studies of cell-type dependent connectivity.

\subsection{Inference} 
We perform posterior inference via Markov-chain Monte Carlo (MCMC),
annealing on the global likelihood during the traditional burn-in
phase. MCMC transition kernels for different parts of the state space
can be chained together to construct a kernel whose ergodic
distribution is the target ergodic distribution over the entire state space.

Our first transition kernel (``structural'') performs gibbs sampling 
of the assignment vector $p(\vec{c} | \eta, \theta, \alpha)$. 
The lack of conjugacy in our likelihood model makes an explicit 
evaluation of the conditional assignment probabilities impossible, 
motivating us to use an auxiliary variable method \autocite{Neal2000}
in which a collection of ephemeral classs are explicitly represented
for the duration of the Gibbs scan.

We then employ a transition kernel to update the per-component
parameter values $\eta_{mn}$. Conditioned on the assignment vector
$\vec{c}$ and the model hyperparameters $\theta, \alpha$ the 
individual $\eta_{mn}$ are independent. We slice sample \autocite{Neal2003}
each component's parameters, choosing the slice width as a function
of the global hyperparameter range.

The global hyper-parameters, both $\alpha$ and $\theta$, are allowed
to take on a discrete set of possible values. As $\theta$ is often a
tuple of possible values, we explore the cartesian product of all
possible values. We then Gibbs sample (our final transition kernel),
which is always possible in a small, finite, discrete state space.

We chain these three kernels together, and then globally anneal on the
likelihood from a temperature of $T=64$ down to $T=1$ over $900$
iterations unless otherwise indicated, and then run the chain for
another $100$ iterations. We then generate at least $20$ samples, each
taken from the end of a single Markov chain initialized from different
random initial points in the state space. For visualization we pick the
chain with the highest log likelihood, but for all numerical
comparisons (including link probability and ARI) we use this full
collection of samples from the posterior distribution to estimate the
resulting statistics.

\subsection{Link Prediction}
As a proxy for link-prediction accuracy we compute the probability of a link
between two cells using each model, trained fully on the data. While this method
is potentially prone to overfitting, the overfitting will be shared across models
and in fact will preferentially bias in favor of competing models
which over-cluster the data. We use a full collection of posterior samples when
computing the link probability, and then compute the area under the ROC curve for
each.

\subsection{Model Comparison}
We compare our model with a standard network clustering model, the
latent-position clutsering model. This model assumes each cell belongs
to one of K clusters, and each cluster is associated with a
$d-$dimensional Gaussian distribution. The probability of a link is
then a function of the distance between the data points in this
continuous-space. We use \autocite{VBLPCM} a variational
implementation provided in R, parametrically varying the number of
latent dimensions and the number of requested groups.  While this
model provides reasonable link predictive accuracy, the clusterings
dramatically disagree with those from human anatomists.

\subsection{Parameters}

Hierarchical generative models can be sensitive to hyperparameter
settings, thus for most hyperparamters we perform inference. In cases
where we cannot we run separate collections of markov chains at
separate settings and show the results across all pooled
parameters. For the case of the mouse retina data, we consisder
maximum link probability $p_{max} \in \{0.95, 0.9, 0.7\}$, variance
scales for the synapse density profile of $\sigma^2 \in \{0.01, 0.1,
1.0\}$ (of normalized depth), and $K \in \{2, 3\}$ possible synapse
density profile mixture components. For the connectivity-distance-only
model we actually perform inference over both $p_{max}$ and $p_{min}$.

\subsubsection{Mixing of our Markov chains} 
Evaluating whether or not approximate inference methods, such as MCMC,
produce samples which are valid approximations of the posterior
distribution is an ongoing area of research in the computational
statistics community. We use a rough proxy here -- synthetic
likelihood evaluation.  For synthetic datasets of sizes comparable to
our real data size, do we recover known ground truth information after
running our markov chains for the appropriate amount of time?

Figures \ref{fig:mixing:ari} and \ref{fig:mixing:scorevstime} shows
the ARI to ground truth and the total log score as a function of
runtime.  We see dramatic changes in log score initially as we vary
the temperature, stabilizing as runtime progresses, for each
chain. Then we see the characteristic jumps between nearby modes
towards the end of the run, in both log score and ARI.  Importantly,
regardless of whether our model over- or under-estimates the exact
posterior variance about the network, we find points in the latent
variable space that are both predictive \textit{and} parsimonious,
largely agreeing with the human anatomists and predicting existing
connections.

>>>>

\subsection*{Types and connectivity in artificial structures}
\begin{figure}
  \centering 
  \centerline{\includegraphics[width=4in]{mos6502.ai}}
  \caption{Discovering connectivity and type the 6502 microprocessor.
    a.) is the micrograph of the original microprocessor, with the
    region containing the registers under study highlighted. b.) Our
    graph consists of the interconnections of MOS field-effect
    transistors with three terminals, Gate, C1, and C2. Reconstruction
    technique did not permit resolution of C1 and C2 into source and
    drain. c.) The spatial distribution of the transistors in each
    cluster show a clear pattern d.) The clusters and connectivity
    versus distance for connections between Gate and C1, Gate and C2,
    and C1 and C2 terminals on a transistor. Red and teal types have a
    terminal pulled down to ground and mostly function as
    inverters. Purple class are clocked, stateful transistors, orange
    control the ALU and yellow control the special data bus (SDB).}
  \label{fig:mos6502}
\end{figure}

To show the applicability of our method to other connectome-style
datasets, we obtained the spatial location and interconnectivity of
the transistors in a classic microprocessor, the MOS Technology 6502
(used in the Apple II) \autocite{James2010}. Computer architects use
common patterns of transistors when designing circuits, with each
transistor having a ``type'' in the circuit. We identified a region of
the processor with complex but known structure containing the primary
8-bit registers X, Y, and S (fig~\ref{fig:mos6502}).

Our algorithm identifies areas of spatial homogeneity that mirror the
known structure in the underlying architectural circuit, segmenting
transistor types recognizable to computer architects. Using the
original schematics, we see that one identified type contains the
``clocked'' transistors, which retain digital state. Two other types
contain transistors with pins C1 or C2 connected to ground, mostly
serving as inverters.  An additional identified type controls the
behavior of the three registers of interest (X, Y, and S) with respect
to the SB data bus, either allowing them to latch or drive data from
the bus. The repeat patterns of spatial connectivity are visible in
figure~\ref{fig:mos6502}c, showing the man-made horizontal and
vertical layout of the same types of transistors.
